---
title: "Assignment 1 - Language Development in ASD - part 4"
author: "Riccardo Fusaroli"
date: "August 10, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Welcome to the fourth exciting part of the Language Development in ASD exercise

In this exercise we will assess how many participants we would need to adequately replicate our findings (ensuring our sample size is adequate, our alpha at 0.05 and our beta at 0.8).

### Exercise 1

How much power does your study have (if your model estimates are quite right)?
- [GitHub]Load your dataset, fit your favorite model, assess power for your main effects and interactions of interest.
- Report the power analysis and comment on what you can (or cannot) use its estimates for.




ANSWER:

I tested the power of the study for finding the effect sizes given by the model.
The power for all the effects is assessed to be at least above 0.91, far more than the 0.8 threshold.
But this is not necessarily a reliable assessment. The observed effect sizes are unreliable if an experiment is underpowered, meaning that the observed effect sizes could be higher than the actual effect sizes. This means that the power estimates might also be exaggerated. Therefore, the minimum interesting effect sizes should instead be used to calculate the power of the study.

```{r}
#Libraries
library(simr)
library(lmerTest)
library(MASS)

#Read datasets from before
d = read.csv("Assignment1CLEAN.csv", sep = ",", stringsAsFactors = F, header = T)
dt = read.csv("Assignment1TestingCLEAN.csv", sep = ",", stringsAsFactors = F, header = T)

#Change Subject ID's to not overlap
dt$SUBJ = dt$SUBJ+61

#Combine them
d = rbind(d, dt)

#Make variables to factors
d$SUBJ=as.factor(d$SUBJ)
d$Diagnosis=as.factor(d$Diagnosis)

#Recreate chosen model
Model_Final = lmer(CHI_MLU ~ VISIT*ADOS + MOT_MLU + types_CHI + verbalIQ + (1+VISIT|SUBJ), d)

summary(Model_Final)

#Test power with current beta-values - for each variable
powerVISIT = powerSim(Model_Final, fixed("VISIT"), nsim=100)
powerADOS = powerSim(Model_Final, fixed("ADOS"), nsim=100)
powerMOT_MLU = powerSim(Model_Final, fixed("MOT_MLU"), nsim=100)
powertypes_CHI = powerSim(Model_Final, fixed("types_CHI"), nsim=100)
powerverbalIQ = powerSim(Model_Final, fixed("verbalIQ"), nsim=100)
powerVISIT_ADOS = powerSim(Model_Final, fixed("VISIT:ADOS"), nsim=100)

powerVISIT
powerADOS
powerVISIT_ADOS
powertypes_CHI
powerMOT_MLU
powerverbalIQ

```

### Exercise 2

How would you perform a more conservative power analysis?
- Identify and justify a minimum effect size for each of your relevant effects
- [GitHub] take the model from exercise 1 and replace the effects with the minimum effect size that you'd accept.
- [GitHub] assess the power curve by Child.ID, identifying an ideal number of participants to estimate each effect
- OPTIONAL if your power estimates do not reach an acceptable threshold simulate additional participants and repeat the previous analysis
- Report the power analysis and comment on what you can (or cannot) use its estimates for.







Answer:

For the minimum interesting effect sizes, based very conservatively on the observed effect sizes, I have chosen the following:
Visit: 0.05
ADOS: 0.01
Mother's MLU: 0.05
Word types: 0.003
Verbal IQ: 0.01
Visit and ADOS interaction:-0.005

INSERT PLOT

The plot shows that there is not enough participants to have a power of 0.8. But there are not enough participants in the dataset to find out how many participants would be needed as a minimum.

Therefore, I will simulate new participants. I will use Riccardo's function for that, which can omly generate values for visit and diagnosis. Thus, based on running a basic model like that, I choose minimum values of
Visit: 0.05
Diagnosis -0.1
Visit and diagnosis interaction: 0.1

```{r}
# Change to lowest relevant effect sizes
Model_Final2 = Model_Final
fixef(Model_Final2)
fixef(Model_Final2)["VISIT"] <- 0.05
fixef(Model_Final2)["ADOS"] <- 0.01
fixef(Model_Final2)["MOT_MLU"] <- 0.05
fixef(Model_Final2)["types_CHI"] <- 0.003
fixef(Model_Final2)["verbalIQ"] <- 0.01
fixef(Model_Final2)["VISIT:ADOS"] <- -0.005


#Calculate the power depending on number of participants
powerCurveVISIT = powerCurve(Model_Final2, fixed("VISIT"), along="SUBJ", nsim=100)
powerCurveADOS = powerCurve(Model_Final2, fixed("ADOS"), along="SUBJ", nsim=100)
powerCurveVISIT_ADOS = powerCurve(Model_Final2, fixed("VISIT:ADOS"), along="SUBJ", nsim=100)
powerCurvetypes_CHI = powerCurve(Model_Final2, fixed("types_CHI"), along="SUBJ", nsim=100)
powerCurveMOT_MLU = powerCurve(Model_Final2, fixed("MOT_MLU"), along="SUBJ", nsim=100)
powerCurveverbalIQ = powerCurve(Model_Final2, fixed("verbalIQ"), along="SUBJ", nsim=100)

#And Plot it
plot(powerCurveVISIT)
plot(powerCurveADOS)
plot(powerCurveVISIT_ADOS)
plot(powerCurvetypes_CHI)
plot(powerCurveMOT_MLU)
plot(powerCurveverbalIQ)


```

```{r}
#RICCARDO's FUNCTION for simulating new participants:
#edited his function a little to fit with my variable names

#############

### Riccardo's clumsy function to simulate new participants
### TO DO points are only notes for myself, so not part of the assignment


createNewData <- function (participants,visits,model){
  # participants is the number of subjects
  # visits is the number of visits
  # TO DO: LOOP THROUGH ALL FE ROWS AND AUTOMATICALLY EXTRACT NAMES OF FIXED EFFECTS AND ESTIMATES
  fe <- fixef(model)
  Intercept <- fe[1] #intercept
  bVisit <- fe[2] #visit
  bDiagnosis <- fe[3] #diagnosis
  bVisitDiagnosis <- fe[4] #visit diagnosis interaction
  # TO DO: INTEGRATE STANDARD ERROR?
  
  # TO DO: LOOP THROUGH ALL VC COMPONENTS AND AUTOMATICALLY EXTRACT NAMES OF EFFECTS AND ESTIMATES
  vc<-VarCorr(model) # variance component
  sigmaSubject <- as.numeric(attr(vc[[1]],"stddev")[1]) # random intercept by subject
  sigmaVisit <- as.numeric(attr(vc[[1]],"stddev")[2]) # random slope of visit over subject
  sigmaResiduals <- as.numeric(attr(vc,"sc"))
  sigmaCorrelation <- as.numeric(attr(vc[[1]],"correlation")[2])
  
  # Create an empty dataframe
  d=expand.grid(VISIT=1:visits,SUBJ=1:participants)
  # Randomly sample from a binomial (to generate the diagnosis)
  condition <- sample(rep(0:1, participants/2))
  d$Diagnosis<-condition[d$SUBJ]
  d$Diagnosis[is.na(d$Diagnosis)]<-1
  
  ## Define variance covariance matrices:
  Sigma.u<-matrix(c(sigmaSubject^2,
                    sigmaCorrelation*sigmaSubject*sigmaVisit,
                    sigmaCorrelation*sigmaSubject*sigmaVisit,
                    sigmaVisit^2),nrow=2)
  
  ## generate new fake participants (column1=RandomIntercept, column2=RandomSlope)
  u<-mvrnorm(n=participants,
             mu=c(0,0),Sigma=cov(ranef(model)$SUBJ))
  
  ## now generate fake data:
  ### the outcome is extracted from a gaussian with
  ### the solution to the model's equation as mean and
  ### the residual standard deviation as standard deviation 
  d$CHI_MLU <- rnorm(participants*visits,
                     (Intercept+u[,1]) +
                     (bVisit+u[,2])*d$VISIT + 
                     bDiagnosis*d$Diagnosis ,sigmaResiduals)  
  
  return(d)
}


```

```{r}
#Simulate dataset
d_sim = createNewData(500, 6, Model_Final)
d_sim$SUBJ=as.factor(d_sim$SUBJ)

#Train model on simulated data
Model_Sim = lmer(CHI_MLU ~ VISIT*Diagnosis + (1+VISIT|SUBJ), d_sim)
summary(Model_Sim)


#Run a basic model to find smallest interesting effect for the model on the simulated data
Model_0 = lmer(CHI_MLU ~ VISIT*Diagnosis + (1+VISIT|SUBJ), d)
summary(Model_0)


#Define smallest interesting effect size
fixef(Model_Sim)["VISIT"] <- 0.05
fixef(Model_Sim)["Diagnosis"] <- -0.1
fixef(Model_Sim)["VISIT:Diagnosis"] <- 0.1


#Calculate power for different amounts of participants
powerCurveVISITsim = powerCurve(Model_Sim, fixed("VISIT"), along="SUBJ", nsim=100)
powerCurveDiagnosissim = powerCurve(Model_Sim, fixed("Diagnosis"), along="SUBJ", nsim=100)
powerCurveinteractsim = powerCurve(Model_Sim, fixed("VISIT:Diagnosis"), along="SUBJ", nsim=100)

plot(powerCurveVISITsim)
plot(powerCurveDiagnosissim)
plot(powerCurveinteractsim)

```


### Exercise 3

Assume you have only the resources to collect 30 kids (15 with ASD and 15 TDs). Identify the power for each relevant effect and discuss whether it's worth to run the study and why.

Answer:


I look at the powerplots, and see if 30 participants are enough for anything.
I don't think it is.


```{r}

plot(powerCurveVISIT)
plot(powerCurveADOS)
plot(powerCurveVISIT_ADOS)
plot(powerCurvetypes_CHI)
plot(powerCurveMOT_MLU)
plot(powerCurveverbalIQ)

```


